---
title: "DiFrank_HW2"
output:
  word_document: default
  html_document: default
date: "2025-01-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

Assumptions: The data is drawn from a multivariate normal distribution. Observations are independent and drawn randomly. The s is a good estimate of population covariance matrix.

```{r}
# Given data
x_bar <- c(5.936, 2.770, 4.260, 1.326)
s <- matrix(c(0.40, 0.09, 0.30, 0.05,
              0.09, 0.10, 0.07, 0.05,
              0.30, 0.07, 0.30, 0.05,
              0.05, 0.05, 0.05, 0.08), nrow = 4, byrow = TRUE)
n <- 50

#Compute variance of (x_bar1 - x_bar2)
var_diff <- s[1, 1] + s[2, 2] - 2 * s[1, 2]

#Compute T^2
T_squared <- n * (x_bar[1] - x_bar[2])^2 / var_diff

#Compute critical value
alpha <- 0.05
critical_value <- qf(1 - alpha, df1 = 1, df2 = n - 1)

#Output results
list(T_squared = T_squared, Critical_Value = critical_value, Reject_H0 = T_squared > critical_value)
```

At alpha = 0.05, we have statistically significant evidence to suggest we reject the null hypothesis that u1 =u2.

## Problem 2 

**Part (a) We test if there is a difference among the mean reaction times across the four treatments.**

#### Hypotheses:

-   Null hypothesis (Ho​): u1=u2=u3=u4, i.e. no overall treatment effect.

-   Alternative hypothesis (Ha): At least one u1​ differs; there is an overall treatment effect.

We assume the data is drawn from a p=variable normal distribution and observations are random and independent; there is no carry-over effect between treatments.

```{r}
dat <- read.table("data/T6-8.DAT", header = FALSE)
colnames(dat) <- c("trt1", "trt2", "trt3", "trt4") 
head(dat)
```

Perfoming Hotelling's T\^2 hypothesis test:

```{r}
library(ICSNP)
mu0=rep(0,ncol(dat))
HotellingsT2(X=dat,mu=mu0)
```

According to our test, we have significant evidence (p value \<2.2e-16) to reject the null hypothesis that there is no overall treatment effect (the mean vector is not 0). At least one of the four measures is a non-zero mean difference before and after treatment.

**Part (b) Construct 95% simultaneous CIs for contrasts representing the treatments mean difference u1-u2, u1-u3, and u1-u3. Interpret the results.**

Function for simultaneous CIs:

```{r}
simul.CI <- function(data, level = 0.95){
  xbar <- colMeans(data)
  S<-cov(data)
  n<-nrow(data)
  p<-ncol(data)
  crit<- (n-1)*p/((n-p)*n) * qf(1-level,p,n-p,lower.tail = FALSE)
  H<-sqrt(crit*diag(S))
  out<-data.frame(Estimate = xbar, Lower = xbar-H, Upper =xbar+H)
  return(out)
}
```

Doing simultaneous CIs:

```{r}
sint<- simul.CI(dat)
round(sint,3)
```

Because 0 is not included in the confidence intervals, there is strong evidence that the true mean treatment difference is not 0 at 95% confidence.

**Part (c)  Repeat (a) by using a contrast matrix that describes interaction effect, and each of the two main effects.**

We assume there is no carry over affect between the treatments. We assume the data follows a normal distribution and was randomly sampled.

Function for testing multiple contrasts:

```{r}
T2.contrast<- function(data.matrix, contrast.matrix,alpha = 0.05){
  Xmat<- data.matrix
  C<- contrast.matrix
  n<- nrow(Xmat)
  q<- nrow(C)
  xbar<- colMeans(Xmat)
  S<- cov(Xmat)
  #intermediate quantities
  invCSC<- solve(C%*%S %*% (t(C)))
  Cxbar<- C %*% xbar
  #test statistic
  T2<- n*(n-q)/((n-1)*q) * (t(Cxbar)) %*% invCSC %*% (Cxbar)
  #critical value
  critical_F = qf(alpha, df1=q,df2=n-q,lower.tail = FALSE)
  #p value
  pv<- pf(T2, df1=q, df2=n-q,lower.tail = FALSE)
  #display results
  results <- data.frame(T2=T2, Fcritical=critical_F,df1=q,df2=n-q,pvalue=pv)
  return(results)
}
```

Contrast matrix and performing the test:

```{r}
C<- cbind(c(1,1,1),-diag(1,3))
C
#test 
T2.contrast(dat, C)
```

There is strong evidence at alpha = 0.05 that the four treatment means are different (i.e. strong evidence to reject the null hypothesis that the 4 means are equal).

## Problem 3 

```{r}
#reading in data 
dat3<- read.table("data/T1-8.DAT", header = FALSE)
colnames(dat3) <- c("DominantRadius", "Radius", "DominantHumerus" ,"Humerous","DominantUlna","Ulna")
head(dat3)
```

**Part (a) For each bone, define the difference response between the dominant and non-dominant side. If the true mean difference is zero, it means that the mean mineral content is the same for dominant and non-dominant bones. We are interesting in formally testing whether the mean difference is different than zero.**

μD​=[μD1​​,μD2​​,μD3​​] is the vector of mean differences (Di for the three bones)

Null hypothesis: Ho: μD = 0 (the mean differences are all zero across bones)

Alternative hypothesis: Ha: μD /= 0 (at least one mean difference is significantly different from zero)

Test name: Hotelling's T2 Test, where the sample mean vector is Dˉ, the sample covariance matrix S of the differences.

**Part (b) Performance of the test**

```{r}
mu0=rep(0,ncol(dat3))
HotellingsT2(X=dat3,mu=mu0)
```

At alpha=0.05, there is statistically significant evidence to reject the null (i.e. the true mean difference is not equal to 0).

**Part (c) Bonferroni 95% intervals for each component of the difference of the two mean vectors**

```{r}
#differences
diff_rad <- dat3$DominantRadius - dat3$Radius
diff_hum <- dat3$DominantHumerus - dat3$Humerous
diff_ulna <- dat3$DominantUlna - dat3$Ulna

#mean differences
mean_diff_rad <- mean(diff_rad)
mean_diff_hum <- mean(diff_hum)
mean_diff_ulna <- mean(diff_ulna)

#standard errors
se_rad <- sd(diff_rad) / sqrt(length(diff_rad))
se_hum <- sd(diff_hum) / sqrt(length(diff_hum))
se_ulna <- sd(diff_ulna) / sqrt(length(diff_ulna))

#bonferroni adjustment
alpha = 0.05
k <- 3 #number of tests
adjusted_alpha <- alpha / k
t_critical <- qt(1 - adjusted_alpha / 2, df = length(diff_rad) - 1)

#CIs 
ci_rad <- c(
    mean_diff_rad - t_critical * se_rad,
  mean_diff_rad + t_critical * se_rad
)

ci_hum <- c(
    mean_diff_hum - t_critical * se_hum,
  mean_diff_hum + t_critical * se_hum
)

ci_ulna <- c(
    mean_diff_ulna - t_critical * se_ulna,
  mean_diff_ulna + t_critical * se_ulna
)

#results 
cat("Bonferroni 95% Confidence Intervals:\n")
cat("Radius:", ci_rad, "\n")
cat("Humerus:", ci_hum, "\n")
cat("Ulna:", ci_ulna, "\n")
```

For the radius, ulna, and humerus, the intervals include zero, indicating there is not statistically significant evidence to reject a null of zero difference between dominant and non-dominant mineral content.

## Problem 4 

```{r}
#reading in data 
dat4<- read.table(file="data/middleschool.txt", sep=",",header=TRUE)
colnames(dat4)
head(dat4)
```

**Part (a) Construct Bonferroni intervals for the mean performance in each of the four subjects, MATH, ELA, Science and Social studies, for school A.**

```{r}
#Filter data for School A and School B
school_A <- dat4[dat4$school == "A", ]
school_B <- dat4[dat4$school == "B", ]

#Function to calculate Bonferroni intervals
bonferroni_intervals <- function(data, subjects, alpha = 0.05) {
  k <- length(subjects)  # Number of comparisons
  adjusted_alpha <- alpha / k
  t_critical <- qt(1 - adjusted_alpha / 2, df = nrow(data) - 1)
  
  results <- data.frame(
    Subject = subjects,
    Mean = sapply(subjects, function(subj) mean(data[[subj]], na.rm = TRUE)),
    Lower = NA,
    Upper = NA
  )
  
  results$SE <- sapply(subjects, function(subj) {
    sd(data[[subj]], na.rm = TRUE) / sqrt(nrow(data))
  })
  
  results$Lower <- results$Mean - t_critical * results$SE
  results$Upper <- results$Mean + t_critical * results$SE
  return(results)
}

# Subjects to analyze
subjects <- c("Math", "ELA", "Science", "SocialStudies")

# Bonferroni intervals for School A
bonferroni_A <- bonferroni_intervals(school_A, subjects)
bonferroni_A
```

**Part (b) Repeat (a) for school B.**

```{r}
bonferroni_B <- bonferroni_intervals(school_B, subjects)
bonferroni_B
```

**Part (c) Use (a) and (b) and comment on whether any school has better performance in any subject.**

For Math, ELA, Science, and Social Studies, the confidence intervals between each school overlap, so I wouldn't make any strong conclusions on the differences between schools. Even though the mean is lower in each subject for school A, we are using 95% confidence, and our 95% confidence interval for each of these subjects in A contains the mean for school B.
