---
title: "in class"
output: html_document
date: "2025-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 01-16-25

## Standard normal distributions

Verifying- example with in class R commands

```{r}
#default is standard normal if you dont specify mean and sd
qnorm(0.025)
qnorm(0.975)
```

As shown, within 2 sd of the mean we should observe around 95% of the data (0.025-0.975)

# 01/21/25

```{r}
#install.packages("ICSNP")
library("ICSNP")

data("LASERI")
head(LASERI)
#we want to compare different indicators- cardiac output before and after tilt therapy
#focusing on 4 different measures for this - 
laseri <- LASERI[,25:28]

#looking at pair plot will show how these four variables relate to eachother
pairs(laseri)

```

what is the true mean difference between the pre and post tilt for each of the four measures of differences? can we formally assess whether there is a change between pre and post for all these measures? is the tilting experiment having a healthy impact?

```{r}

```

## 1/23/25

univariate hyp test- manual way-

```{r}
mu0 <- 0 #null value 

n<- nrow(LASERI)
alpha <- 0.05

#t crit value using built in r function 
z.crit <- qnorm(p=alpha/2,lower.tail=F)

xbar =mean(laseri$HRT1T4)
stddev = sd(laseri$HRT1T4)
test.stat <- (xbar-mu0)/(stddev/sqrt(n))
pval<- 2*pnorm(abs(test.stat),lower.tail = FALSE)
pval #significant 
```

built in t test function -

```{r}
t.test(laseri$HRT1T4)
```

-   note if the df (n-1) wasnt large enough, you would need to check normality. here it is large enough

-   note 95% CIs are given automatically here as well
